scRNAseq benchmark in want of a title

### 2019-05-27 bmc_article.tex
Updates from Overleaf

We next investigated the impact of different normalization strategies. In addition to log-normalization, the standard Seurat clustering pipeline scales features so that the PCA is not dominated by a small subset of the features, and we therefore included versions of the different normalization, with or without a subsequent scaling (except sctransform's variance-stabilizing transformation, given that, conceptually, it involves something analogous to scaling). Seurat's scaling function includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features. Finally, it has been proposed that the use of stable genes, and in particular cytosolic ribosomal genes, can be used to normalized scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features, removing the confounding covariation with the subpopulation (Supplementary Figure 12A). The simple step of scaling tended to remove much of the correlation with these features, and in the absence of scaling, standard Seurat normalization resulted in fairly high correlation. sctransform had the lowest correlation, but most methods (including normalization based on stable genes) were able to remove most of the correlation when combined with scaling. The only exception is Seurat normalization with scaling regressing out the number of features, which surprisingly increased the correlation.

We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure 13). For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure \ref{fig:figure3}). Regressing out covariates systematically gave poorer performance on all metrics. sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure \ref{fig:figure3}), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13). Of note, even though sctransform is meant to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets. Interestingly, some clusters tended to be less well distinguished with sctransform than with non-scaled (and in some cases non-normalized) methods, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.

@@

A standard clustering pipeline typically involves a step of selecting highly-variable genes, which is complicated by the digital nature and the mean-variance relationship of RNAseq. Seurat's earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions rely of a different measure of variance, again standardized. Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this bias reflects biological relevance and hence is actually helpful in classifying cell types. Another common practice for selecting features is to use those with the highest mean expression, and recently \citep{townesGlmpca2019} instead suggested to use deviance, while sctransform provides its own ordering the genes by transformed variance. We therefore compared these methods.

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.6]{{main_figures/figure4_selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained. \texbf{B:} Accuracy of clustering (at the right number of clusters) based on the given clustering methods and number of genes selected.}
    \label{fig:figure4}
\end{figure}

@@

The main observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and Seurat's un-standardized estimate of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering.

>>>>

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in Seurat, we tested scran's pooling-based normalization \citep{lunPooling2016}, sctransform's variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}. In addition to log-normalization, the standard Seurat clustering pipeline scales features so that the PCA is not dominated by a small subset of the features, and we therefore included versions of the different normalization, with or without a subsequent scaling (except sctransform's variance-stabilizing transformation, given that, conceptually, it involves something analogous to scaling). Finally, Seurat's `ScaleData` includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features.

An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an imperfect correlation \citep{hafemeisterSCtransform2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features (Supplementary Figure 12A). In the 10X datasets, most methods successfully decreased the correlation with covariates, including the mere scaling of non-normalized data. This was not the case in the smartseq datasets, which required the additional regressing out of covariates.

(NOTE: a correlation of PC1 with library size might be caused by the fact that subpopulations have a different average library size. I'm repeating this evaluation with the variance explained by libSize in a PC1\textasciitilde subpopulation+libSize model.)

We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure 13). For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure \ref{fig:figure3}). Regressing out covariates systematically gave poorer performance on all metrics. sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure \ref{fig:figure3}), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13). Of note, even though sctransform is meant to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets. Interestingly, clusters with the lowest average silhouette width (and which are therefore more difficult to classify) tended to be less well distinguished with sctransform, and instead showed a good difference in non-normalized and/or non-scaled data, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.

@@

A standard clustering pipeline typically involves a step of selecting highly-variable genes, which is complicated by the digital nature and the mean-variance relationship of RNAseq. Seurat's earlier approaches involved the use of dispersion estimates standardized for the expression levels, while more recent versions rely of a different measure of variance, again standardized. Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is nevertheless plausible that this bias reflects biological relevance and hence is actually helpful in classifying cell types. \citep{townesGlmpca2019} instead suggested to use deviance, while sctransform provides its own ordering the genes by transformed variance. We therefore compared these methods.

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.6]{{main_figures/figure4_selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. rankMean and rankMax refer to, respectively, the mean rank and the maximum rank in the following metrics: variance, standardized variance, sctransform, deviance. \texbf{B:} Accuracy of clustering (at the right number of clusters) based on the given selection methods and number of genes selected.}
    \label{fig:figure4}
\end{figure}

@@

The main observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and Seurat's unstandardized estimate of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering.
